F\# Data: \textnormal{Making structured data first-class citizens}

### 2015-11-14 pldi-2016/paper.tex
Edits in TP formalization

\documentclass[preprint]{sigplanconf}

\usepackage{amssymb}
\usepackage{multicol}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage[fleqn]{amsmath}
\usepackage{graphics} 
\usepackage{stmaryrd}
\usepackage{amsthm}
\usepackage[hyphens]{url}
\usepackage{hyperref}
\usepackage{ftnright}
\usepackage[T1]{fontenc}
\usepackage{semantic}
\usepackage{enumitem}
%\usepackage[hang,flushmargin]{footmisc} 
%\usepackage[notquote]{hanging}
%\usepackage{flushend}

@@

% TODO: Adapt XML example 
% TODO: Switch from sum types to variant types
% TODO: Discuss 'stability' of the inference (minor change in schema does not cause dramatic cahnges)
% TODO: Say that any variant is a subtype of any other variant 
% TODO: Say that anything is a subtype of variant
% TODO: Say that variant is a last-resort thing
% TODO: Double check that bottom/top are used correctly
% TODO: Explain (even more) that there is no least upper bound, which is why Theorem 3 is good enough
% TODO:  -- but maybe we can have lub w.r.t variants? or something?
% TODO: Avoid "no other rule applies" in the formal system
% TODO: Practical evaluation (lots of downloads, lots of Github - how often people customize samples?)

@@

\newcommand{\lsep}[0]{\;\; | \;\;}
\newcommand{\narrow}[1]{\hspace{-0.85em} #1 \hspace{-0.85em}}

@@

\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}

@@

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}
\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
\copyrightyear{20yy} 
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
\doi{nnnnnnn.nnnnnnn}

\titlebanner{Unpublished draft, March 2015}        % These are ignored unless
%\preprintfooter{short description of paper}   % 'preprint' option specified.

@@

\begin{abstract}
Most static type systems assume that a program runs in a closed world, but this is not the 
case. Modern applications interact with external services and often access data in structured 
formats such as XML, CSV and JSON. Static type systems do not understand such external data sources
and only make accessing data more cumbersome. Should we just give up and leave the messy world of 
external data to dynamic typing and runtime checks?

Of course, we should not give up! In this paper, we show how to integrate external data sources 
into the F\# type system. As most real-world data on the web do not come with an explicit schema, 
we develop a type inference algorithm that infers a type from representative samples. Next, we use 
the type provider mechanism for integrating the inferred structures into the F\# type
system.

The resulting library significantly reduces the amount of code that developers need to write when 
accessing data. It also provides additional safety guarantees. Arguably, as much safety as possible 
if we abandon the incorrect closed world assumption.
\end{abstract}

\category{D.3.3}{Programming Languages}{Language Constructs and Features }
\keywords F\#, Type Providers, JSON, XML, Data, Type Inference

@@

The purpose of many modern applications is to connect multiple data sources or services and
present the aggregated data in a new form. Mobile applications for taking notes, searching train
schedules or finding tomorrow's weather all communicate with one or more services over the network.

Increasing number of such services provide REST-based end-points that return data as CSV, XML
or JSON. Despite numerous schematization efforts, the services typically do not come with an explicit 
schema. At best, the documentation provides a number of typical requests and sample responses.

For example, the OpenWeatherMap service provides an end-point for getting current weather for a 
given city\footnote{See ``Current weather data'': \url{http://openweathermap.org/current}}. The page documents the input 
URL parameters and shows one sample JSON response to illustrate the response structure.
Using a standard library for working with JSON and HTTP, we might call the service and 
read the temperature as follows\footnote{We abbreviate the full URL: 
\url{http://api.openweathermap.org/data/2.5/weather?q=Prague\&units=metric}}:

\noindent
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{data}=\ident{Http.Request}(\str{http://weather.org/?q=Prague}) \\[0.1em]
 \kvd{match}~\ident{JsonValue.Parse}(\ident{data})~\kvd{with} \\[0.1em]
 |~\ident{Record}(\ident{root})\rightarrow \\[0.1em]
 \quad \kvd{match}~\ident{Map.find}~\str{main}~\ident{root}~\kvd{with} \\[0.1em]
 \quad |~\ident{Record}(\ident{main})\rightarrow \\[0.1em]
 \quad \quad \kvd{match}~\ident{Map.find}~\str{temp}~\ident{main}~\kvd{with} \\[0.1em]
 \quad \quad |~\ident{Number}(\ident{num})\rightarrow \ident{printfn}~\str{Lovely \%f degrees!}~\ident{num} \\[0.1em]
 \quad \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\[0.1em]
 \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\[0.1em]
 |~\_\rightarrow \ident{failwith}~\str{Incorrect format} 
\end{array}
\end{equation*}
\vspace{0.1em}

\noindent
The code assumes that the response has a particular format described in the documentation. The
root node must be a record with a \str{main} field, which has to be another record containing
a \str{temp} field with a numerical value. When the format is incorrect, the data access simply fails
with an exception.

Using the JSON type provider from the F\# Data library, we can write code with exactly the 
same functionality in two lines:

\vspace{-0.6em}
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{W} = \ident{JsonProvider}\langl\str{http://weather.org/?q=Prague}\rangl \\[0.1em]
 \ident{printfn}~\str{Lovely \%f degrees!}~(\ident{W.GetSample().Main.Temp})
\end{array}
\end{equation*}
\vspace{0.0em}

\noindent
On the first line, $\ident{JsonProvider}\langl\str{...}\rangl$ invokes a type provider at 
compile-time with the URL as a sample. The type provider infers the structure of the response
and provides a type with a \ident{GetSample} method that returns a parsed JSON with nested
properties \ident{Main.Temp}, returning the temperature as a number.

In the rest of the paper, we give a detailed description of the mechanism and we also discuss
safety properties of the approach. The key novel contributions of this paper are:

\begin{itemize}
\item We present type providers for XML, CSV and JSON (Section~\ref{sec:providers}) 
  that are available in the F\# Data library and we also cover practical aspects of the 
  implementation that contributed to their industrial adoption (Section~\ref{sec:impl}).

\item We describe a predictable type inference algorithm for structured data formats that
  underlies the type providers (Section~\ref{sec:inference}). It is based on finding common 
  supertype of a set of examples.

\item We provide a minimal formal model (Section~\ref{sec:formal}) and use it to prove a
  relativized notion of type safety for our type providers (Section~\ref{sec:safety}). Although 
  we focus on our specific case, the relativized safety property demonstrates a new way of 
  thinking about ML-style type safety that is needed in the context of web.
\end{itemize}

\noindent
Although the F\# Data library \cite{fsharp-data} has been widely adopted and is one of the most downloaded F\# libraries 
it has not yet been presented in an academic form. Additional documentation and source code can 
be found at \url{http://fsharp.github.io/FSharp.Data}.

@@

\section{Structural type inference}
\label{sec:inference}

Our type inference algorithm for structured formats is based on a subtyping relation. When 
inferring the type of a document, we infer the most specific types of individual values (CSV rows,
JSON or XML nodes) and find a common supertype of values in the sample.

In this section, we define the \emph{structural type} $\sigma$, which is a type of structured data.
Note that this type is distinct from the programming language types $\tau$ (the type generates the
latter from the former). Next, we define the subtyping relation on structural types $\sigma$ and 
describe the algorithm for finding a common supertype.

Note that the subtyping relation between structural types $\sigma$ does not map to a subtyping 
relation between F\#  types $\tau$. However, it does hold at runtime, meaning that an operation on 
a supertype can be performed on its subtypes. We make the type provider mechanism and runtime aspects 
precise in Section~\ref{sec:formal} before discussing safety in Section~\ref{sec:safety}.

@@

The grammar below defines \emph{structural type} $\sigma$. We distinguish between \emph{non-nullable types}
that always have a valid value (written as $\hat{\sigma}$) and \emph{nullable types} that encompass missing 
and \kvd{null} values (written as $\sigma$). We write $\nu$ for record field names and $\nu_?$
for record names, which can be empty:
%
\begin{equation*}
\begin{array}{rcl}
 \hat{\sigma} &\narrow{=}& \nu_?\; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \\[0.1em]
                &\narrow{|}& \ident{float} \lsep \ident{decimal} \lsep \ident{int} \lsep \ident{bit} \lsep \ident{bool} \lsep \ident{string} 
 \\[0.6em] 
       \sigma &\narrow{=}& ~\hat{\sigma}~ \lsep \kvd{option}\langl \hat{\sigma} \rangl \lsep [\sigma] \\[0.1em]
              &\narrow{|}& \kvd{var}\langl\sigma_1, \ldots, \sigma_n\rangl \lsep ~\top~ \lsep \kvd{null}
\end{array}
\end{equation*}
%
The non-nullable types include records (consisting of an optional name and zero or more fields with
their types) and primitive types. Records arising from XML documents are always named, while records
used by JSON and CSV providers are always unnamed.

Next, we have three numerical types -- \ident{int} for integers, \ident{decimal} for small high-precision 
decimal numbers and \ident{float} for floating-point numbers (these are related by sub-typing relation as
discussed in the next section). Furthermore, the \ident{bit} type is a type of values $0$ and $1$ and 
makes it possible to infer Boolean in the CSV processing example discussed in Section~\ref{sec:providers-csv}.

Any non-nullable type is also a nullable type, but it can be wrapped in the \kvd{option}
constructor to explicitly permit the \kvd{null} value. These are typically mapped to the standard F\# option 
type. A simple collection type $[\sigma]$ is also nullable and missing values or \kvd{null} are treated as 
empty collection. The type $\kvd{null}$ is inhabited by the $\kvd{null}$ value (using an overloaded but not
ambiguous notation) and $\top$ represents the top type.

Finally, a variant type \kvd{var} is annotated with the statically known types it may represent based
on the sample document. As discussed earlier (Section~\ref{sec:providers-xml}), the variant type forces
the user to handle the case when a value is of a different type than the ones statically known. For
this reason, the variant type also implicitly permits the \kvd{null} value.

@@

@@

To provide a basic intuition, the subtyping relation between structural types is illustrated in
Figure~\ref{fig:subtyping-diagram}. We split the diagram in two parts. The upper part shows 
non-nullable types (with records and primitive types). The lower part shows nullable types with 
\kvd{null}, collections and optional values. We omit links between the two part, but any type
$\hat{\sigma}$ is a subtype of $\hat{\sigma}~\kvd{option}$ (in the diagram, we abbreviate 
$\sigma~\kvd{option}$ as $\sigma?$).

The diagram shows only the basic structure and it does not explain relationships between records
with the same name and between union types. This following definition makes this precise.

@@

\vspace{0.5em}
\noindent\boxed{\emph{Primitives, options, collections}}
\begin{align}
\label{eq:sub-prim}\tag{B1}
\ident{float}\,:>\,\ident{decimal}\,:>\,\ident{int}\,:>\,\ident{bit}\quad&\textnormal{and}\quad\ident{bool}\,:>\,\ident{bit}\\[-0.2em]
\label{eq:sub-null}\tag{B2}
\sigma :> \kvd{null} \quad &(\textnormal{iff}~\sigma \neq \hat{\sigma}) \\[-0.2em]
\label{eq:sub-opt}\tag{B3}
\hat{\sigma}~\kvd{option} :> \hat{\sigma} \quad &(\textnormal{for all}~\hat{\sigma})\\[-0.2em]
\label{eq:sub-top}\tag{B4}
\sigma :> \top \quad &(\textnormal{for all}~\sigma)\\[-0.2em]
\label{eq:sub-col}\tag{B5}
[\sigma_1] :> [\sigma_2] \quad &(\textnormal{if}~\sigma_1 :> \sigma_2)
\end{align}

\noindent\boxed{\emph{Variant types}}
\begin{align}
\label{eq:sub-var-bot}\tag{U1}
\kvd{var}\langl \sigma_1, \ldots, \sigma_n\rangl :>&~ \sigma \\
\label{eq:sub-var-eq}\tag{U2}
\kvd{var}\langl \sigma_1, \ldots, \sigma_n \rangl :>&~ \kvd{var}\langl \sigma_1', \ldots, \sigma_n' \rangl
\end{align}

\noindent\boxed{\emph{Record types}}
\begin{align}
\hspace{-1em}
\label{eq:sub-record1}\tag{R1}
 \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \hspace{-0.1em}:> \hspace{-0.1em}
 \nu_?~\{ \nu_1\!:\!\sigma_1', .., \nu_n\!:\!\sigma_n' \}
 \; (\sigma_i \hspace{-0.25em}:>\hspace{-0.25em} \sigma_i')
\end{align}
\vspace{-1.4em}
\begin{align}
\hspace{-1em}
\label{eq:sub-record2}\tag{R2}
 \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \hspace{-0.1em} :> \hspace{-0.1em}
 \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_m\!:\!\sigma_m \}
 \; (m \hspace{-0.1em}\geq\hspace{-0.1em} n)
\end{align}
\vspace{-1.0em}
\begin{align}
\label{eq:sub-record3}\tag{R3}
\hspace{-1.5em}
\begin{array}{l}
 \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} :> \\
 \nu_?~\{ \nu_{\pi(1)}\!:\!\sigma_{\pi(1)}, .., \nu_{\pi(m)}\!:\!\sigma_{\pi(m)} \}
\end{array} \qquad &(\textnormal{permutation}~\pi)
\end{align}
\vspace{-0.6em}
\begin{align}
\label{eq:sub-record4}\tag{R4}
\hspace{-1.5em}
\begin{array}{l}
 \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n, \nu_{n+1}\!:\!\kvd{null}, .., \nu_{n+m}\!:\!\kvd{null} \} :> \\
 \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \}
\end{array}
\end{align}
\end{definition}

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.75,trim=5mm 5mm 5mm 5mm,clip]{images/hierarchy.pdf} % left bottom right top
\end{center}
\vspace{-0.5em}
\caption{Subtype relation between structural types}
\label{fig:subtyping-diagram}
\vspace{-0.5em}
\end{figure}

\begin{figure*}[t]
\noindent
\begin{equation*}
\hspace{3em}
\inference[(record-1)\;]
  { (\nu_i = \nu'_j \Leftrightarrow (i = j) \wedge (i \leq k))
      \qquad \forall i\in\{ 1 .. k \}.(\sigma_i \tsep \sigma'_i \vdash \sigma''_i) }
  { \begin{array}{l}
    \nu? \; \{ \nu_1 : \sigma_1,  \; \ldots \;, \nu_k : \sigma_k, \; \ldots \;, \nu_n : \tau_n \} \tsep
    \nu? \; \{ \nu'_1 : \sigma'_1, \; \ldots \;, \nu'_k : \sigma'_k, \; \ldots \;, \nu'_m : \tau'_m \} \vdash\\
    \nu? \; \{ \nu_1 : \sigma''_1, \; \ldots \; , \nu_k : \sigma''_k, 
                            \nu_{k+1} : \addopt{\sigma_{k+1}}, \ldots, \nu_n : \addopt{\sigma_n},
                            \nu'_{k+1} : \addopt{\sigma'_{k+1}}, \ldots, \nu'_m : \addopt{\sigma'_m} \}
    \end{array} }
\end{equation*}

% union with something else
\begin{equation*}
\textnormal{\footnotesize{(union-1)}}\;\;
\inference
  {\exists i . \tytagof(\sigma_i) = \tytagof(\sigma) & \sigma \tsep \sigma_i \vdash \sigma_i' & \tytagof(\sigma)\neq\ident{union}}
  {\sigma \tsep (\sigma_1 + \ldots + \sigma_n) \vdash (\sigma_1 + \ldots + \dropopt{\sigma_i'} + \ldots + \sigma_n)}
\quad\quad
\inference
  {\nexists i . \tytagof(\sigma_i) = \tytagof(\sigma) & \tytagof(\sigma)\neq\ident{union}}
  {\sigma \tsep (\sigma_1 + \ldots + \sigma_n) \vdash (\sigma_1 + \ldots + \sigma_n + \dropopt{\sigma})}
\end{equation*}

% two union types
\begin{equation*}
\inference[(union-2)\;]
  { \tytagof(\sigma_i) = \tytagof(\sigma'_i) \quad 
    \sigma_i \tsep \sigma_i' \vdash \sigma''_i \qquad (\forall i\in \{ 1 .. k \}) }
  { \begin{array}{l}
    (\sigma_1  + \ldots + \sigma_k +  \ldots + \sigma_n) \tsep 
    (\sigma'_1 + \ldots + \sigma'_k + \ldots + \sigma'_m) \vdash\\
    (\sigma''_1 + \ldots + \sigma''_k + \sigma_{k+1} + \ldots + \sigma_{n} + \sigma'_{k+1} + \ldots + \sigma'_{m}) 
    \end{array} }
%
\qquad\quad
\inference[(union-3)\;]
  {(\textnormal{no other rule applies})}
  {\sigma_1 \tsep \sigma_2 \vdash \dropopt{\sigma_1} + \dropopt{\sigma_2}}
\end{equation*}

>>>>

\documentclass[10pt,preprint,clearpagebib]{sigplanconf}

\usepackage{listings}
\usepackage{enumitem}
\usepackage[hyphens]{url}
\usepackage[svgnames]{xcolor}
\definecolor{lcl}{RGB}{160,0,128}
\usepackage[colorlinks=true,breaklinks,draft=false]{hyperref}
\hypersetup{urlcolor=lcl,linkcolor=lcl}
\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}

\usepackage{amssymb}
\usepackage{multicol}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage[fleqn]{amsmath}
\usepackage{graphics} 
\usepackage{stmaryrd}
\usepackage{amsthm}
\usepackage{ftnright}
\usepackage[T1]{fontenc}
\usepackage{semantic}
\usepackage{enumitem}
\usepackage[hang,flushmargin]{footmisc} 
\usepackage[notquote]{hanging}
\usepackage{flushend}

@@

% TODO: Adapt XML example 
% TODO: Switch from sum types to variant types
% TODO: Discuss 'stability' of the inference (minor change in schema does not cause dramatic cahnges)
% TODO: Say that any variant is a subtype of any other variant 
% TODO: Say that anything is a subtype of variant
% TODO: Say that variant is a last-resort thing
% TODO: Double check that bottom/top are used correctly
% TODO: Explain (even more) that there is no least upper bound, which is why Theorem 3 is good enough
% TODO:  -- but maybe we can have lub w.r.t variants? or something?
% TODO: Avoid "no other rule applies" in the formal system
% TODO: Practical evaluation (lots of downloads, lots of Github - how often people customize samples?)
% TODO: Explain why we need option type (because T + null is a variant)

% TODO: In the PLDI version, we delete all the unnecessary runtime conversions
% but we should say that those happen for practical reasons (e.g. float -> int, or anyhing to string)

% T :> S means that S is a subtype of T (and S can be used where T is expected)
% TOP is the universal supertype
% BOTTOM is the universal subtype

% variant is the top type 
%  (variant is a suprtype of everything; everything is a subtype of variant; 
%   everything can be used where variant is expected)
%
% bottom is the bottom type
%  (bottom is a subtype of everything; everything is a supertype of bottom
%   bottom can be used where anything else is expected - but it has no values)
%
% define :=: when <: and :>, then take equivalence classes

% least upper bound of T1 and T2 is a supertype of both (T :> T1 and T :> T2)
% and for all S such that S :> T1 and S :> T2, it holds that S :> T

@@

\newcommand{\lsep}[0]{\;\; | \;\;}
\newcommand{\narrow}[1]{\hspace{-0.7em} #1 \hspace{-0.7em}}

@@

\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{lemma}[theorem]{Lemma}

@@

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}
\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
\copyrightyear{20yy} 
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
%\doi{nnnnnnn.nnnnnnn}

%\titlebanner{Unpublished draft, March 2015}        % These are ignored unless
%\preprintfooter{short description of paper}   % 'preprint' option specified.

@@

\begin{abstract}
Most modern applications interact with external services and access data in formats such as 
XML, JSON and CSV. Static type systems do not understand external data sources and only 
make accessing data more cumbersome. Should we give up and leave the messy world of external data 
to dynamic typing and runtime checks? Of course, not!

We show how to integrate external data sources into the F\# type system. As most real-world data
do not come with an explicit schema, we develop a type inference algorithm that infers a type from 
representative sample documents and integrate it into the F\# type system using type providers.

Our library significantly reduces the amount of code that developers need to write when 
accessing data. It also provides additional safety guarantees, arguably, as much as possible if 
we abandon the closed world assumption.
\end{abstract}

\category{D.3.3}{Programming Languages}{Language Constructs and Features }
\keywords F\#, Type Providers, Inference, JSON, XML

@@

Applications for taking notes, searching train schedules or finding tomorrow's weather all 
communicate with one or more services over the network and present the aggregated data. 
Increasing number of such services provide REST-based end-points that return data as CSV, XML
or JSON. Despite numerous schematization efforts, most services do not come with an 
explicit schema. At best, the documentation provides sample responses for typical requests.

For example, OpenWeatherMap provides an end-point for getting current weather for a 
given city\footnote{See ``Current weather data'': \url{http://openweathermap.org/current}}. 
The documentation describes the URL parameters and shows one sample JSON to illustrate the typical
response structure. Using a standard library for working with JSON and HTTP, we might call the 
service and read the temperature as follows:
%
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{doc}=\ident{Http.Request}(\str{http://weather.org/?q=NYC}) \\
 \kvd{match}~\ident{JsonValue.Parse}(\ident{doc})~\kvd{with} \\
 |~\ident{Record}(\ident{root})\rightarrow \\
 \quad \kvd{match}~\ident{Map.find}~\str{main}~\ident{root}~\kvd{with} \\
 \quad |~\ident{Record}(\ident{main})\rightarrow \\
 \quad \quad \kvd{match}~\ident{Map.find}~\str{temp}~\ident{main}~\kvd{with} \\
 \quad \quad |~\ident{Number}(\ident{num})\rightarrow \ident{printfn}~\str{Nice \%f degrees!}~\ident{num} \\
 \quad \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
 \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
 |~\_\rightarrow \ident{failwith}~\str{Incorrect format} 
\end{array}
\end{equation*}
%
The code assumes that the response has a particular format described in the documentation. The
root node must be a record with a \str{main} field, which has to be another record containing
a numerical \str{temp} field. When the format is incorrect, the data access simply fails
with an exception.

Using the JSON type provider from F\# Data, we can write code with exactly the 
same functionality in two lines:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{W} = \ident{JsonProvider}\langl\str{http://weather.org/?q=NYC}\rangl \\[0.1em]
 \ident{printfn}~\str{Nice \%f degrees!}~(\ident{W.GetSample().Main.Temp})
\end{array}
\end{equation*}
%
Here, $\ident{JsonProvider}\langl\str{...}\rangl$ invokes a type provider at 
compile-time with the URL as a sample. The type provider infers the structure of the response
and provides a type with a \ident{GetSample} method that returns a parsed JSON with nested
properties \ident{Main.Temp}, returning the temperature as a number.

The rest of the paper describes the mechanism and discusses its safety properties. 
The key novel contributions are:

\begin{itemize}
\item We present F\# Data type providers for XML, CSV and JSON (\S\ref{sec:providers}) 
  and practical aspects of their implementation that contributed to their industrial 
  adoption (\S\ref{sec:impl}).

\item We describe a predictable type inference algorithm for structured data formats that
  underlies the type providers (\S\ref{sec:inference}) based on a \emph{common supertype}
  relation.

\item We give a formal model (\S\ref{sec:formal}) and use it to prove
  \emph{relativized type safety} for the type providers (\S\ref{sec:safety}).
  This adapts the ML-style type safety for the context of the web.
\end{itemize}

% Although the F\# Data library \cite{fsharp-data} has been widely adopted and is one of the most downloaded F\# libraries 
% it has not yet been presented in an academic form. Additional documentation and source code can 
% be found at \url{http://fsharp.github.io/FSharp.Data}.

@@

\newpage
~
\newpage
\section{Structural type inference}
\label{sec:inference}

Our type inference algorithm for structured formats is based on a subtyping relation. When 
inferring the type of a document, we infer the most specific types of individual values (CSV rows,
JSON or XML nodes) and recursively find a common supertype of all child nodes or sample documents.

In this section, we define the \emph{structural type} $\sigma$, which is a type of structured data.
Note that this type is distinct from the programming language types $\tau$ (type providers generate 
the latter from the former). Next, we define the subtyping relation on structural types $\sigma$ and 
describe the algorithm for finding a common supertype.

@@

We distinguish between \emph{non-nullable types} that always have a valid value (written as 
$\hat{\sigma}$) and \emph{nullable types} that encompass missing and \kvd{null} values 
(written as $\sigma$). We write $\nu$ for record field names and $\nu_?$ for names that can be empty:
%
\begin{equation*}
\begin{array}{rcl}
 \hat{\sigma} &\narrow{=}& \nu_?\; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \\[0.1em]
                &\narrow{|}& \ident{float} \lsep \ident{decimal} \lsep \ident{int} \lsep \ident{bool} \lsep \ident{string} 
 \\[0.6em] 
       \sigma &\narrow{=}& ~\hat{\sigma}~ \lsep \kvd{option}\langl \hat{\sigma} \rangl \lsep [\sigma] \\[0.1em]
              &\narrow{|}& \kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl \lsep ~\bot~ \lsep \kvd{null}
\end{array}
\end{equation*}
%
Non-nullable types include records (consisting of an optional name and fields with their types) 
and primitive types. Records arising from XML documents are named, while records used by 
JSON and CSV providers are unnamed.

The three numerical types, \ident{int} for integers, \ident{decimal} for small precise decimal 
numbers and \ident{float} for floating-point numbers, are related by sub-typing as discussed in 
\S\ref{sec:inference-subtyping}.

Any non-nullable type can be wrapped in the \kvd{option} constructor to explicitly permit the 
\kvd{null} value. Type providers map \kvd{option} to standard F\# option type. A collection 
type $[\sigma]$ is also nullable and missing values or \kvd{null} are treated as empty collection. 
The type $\kvd{null}$ is inhabited by the $\kvd{null}$ value (using an overloaded notation) and 
$\bot$ is the bottom type.

A variant type \kvd{any} is annotated with the types it may represent in the sample document.
As discussed earlier (\S\ref{sec:providers-xml}), variants force the user to handle the case when 
a value is of a different type than the statically known ones and so it can be seen as a family of 
top types. For the same reason, the variants also implicitly permit the \kvd{null} value.

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.80,trim=5mm 5mm 5mm 5mm,clip]{images/hierarchy.pdf} % left bottom right top
\end{center}
\vspace{-0.5em}
\caption{Subtype relation between structural types}
\label{fig:subtyping-diagram}
\vspace{-0.5em}
\end{figure}

@@

Figure~\ref{fig:subtyping-diagram} provides a basic intuition about the subtyping between
structural types. The upper part shows non-nullable types (with records and primitive types) and 
the lower part shows nullable types with \kvd{null}, collections and optional values. We omit links 
between the two part, but any type $\hat{\sigma}$ is a subtype of $\kvd{option}\langl\hat{\sigma}\rangl$ 
and we abbreviate $\kvd{option}\langl\sigma\rangl$ as $\sigma?$.

@@

\vspace{0.5em}
\noindent\boxed{\emph{Primitives, options, collections}}
\begin{align}
\label{eq:sub-prim}\tag{B1}
\ident{float}\,&:>\,\ident{decimal}\,:>\,\ident{int}&\\[-0.2em]
\label{eq:sub-null}\tag{B2}
\sigma &:> \kvd{null}  &(\textnormal{iff}~\sigma \neq \hat{\sigma}) \\[-0.2em]
\label{eq:sub-bot}\tag{B3}
\sigma &:> \bot  &(\textnormal{for all}~\sigma)\\[-0.2em]
\label{eq:sub-opt}\tag{B4}
\kvd{option}\langl\hat{\sigma}\rangl &:> \hat{\sigma}  &(\textnormal{for all}~\hat{\sigma})\\[-0.2em]
\label{eq:sub-opt-cov}\tag{B5}
\kvd{option}\langl\hat{\sigma_1}\rangl &:> 
  \kvd{option}\langl\hat{\sigma_2}\rangl  &(\textnormal{if}~\hat{\sigma_1} :> \hat{\sigma_2})\\[-0.2em]
\label{eq:sub-col}\tag{B6}
[\sigma_1] &:> [\sigma_2]  &(\textnormal{if}~\sigma_1 :> \sigma_2)
\end{align}

\noindent\boxed{\emph{Variant types}}
\begin{align}
\label{eq:sub-var-bot}\tag{U1}
\kvd{any}\langl \sigma_1, \ldots, \sigma_n\rangl :>&~ \sigma \\
\label{eq:sub-var-eq}\tag{U2}
\kvd{any}\langl \sigma_1, \ldots, \sigma_n \rangl :>&~ \kvd{any}\langl \sigma_1', \ldots, \sigma_n' \rangl
\end{align}

\noindent\boxed{\emph{Record types}}
\begin{align}
\label{eq:sub-record1}\tag{R1}
\begin{array}{l}
 \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} :> \\
 \nu_?~\{ \nu_1\!:\!\sigma_1', .., \nu_n\!:\!\sigma_n' \}
\end{array} \qquad ~~~~(\textnormal{if}~\sigma_i :> \sigma_i')
\end{align}
\vspace{-0.8em}
\begin{align}
\label{eq:sub-record2}\tag{R2}
\begin{array}{l}
 \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} :> \\
 \nu_?~\{ \nu_1\!:\!\sigma_1, .., .., \nu_m\!:\!\sigma_m \}
\end{array} \qquad (\textnormal{when}~m \geq n)
\end{align}
\vspace{-0.8em}
\begin{align}
\label{eq:sub-record3}\tag{R3}
\begin{array}{l}
 \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} :> \\
 \nu_?~\{ \nu_{\pi(1)}\!:\!\sigma_{\pi(1)}, .., \nu_{\pi(m)}\!:\!\sigma_{\pi(m)} \}
\end{array} ~~(\pi~\textnormal{perm.})
\end{align}
\vspace{-0.8em}
\begin{align}
\label{eq:sub-record4}\tag{R4}
\begin{array}{l}
 \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n, \nu_{n+1}\!:\!\kvd{null} \} :> \\
 \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \}
\end{array}
\end{align}
\end{definition}

@@

\begin{figure*}[t]
\noindent
\begin{equation*}
\boxed{
\begin{array}{c}
\begin{array}{rclcl}
 \tytag &\narrow{=}& \ident{bool} \\
        &\narrow{\lsep}& \ident{string} &\narrow{\lsep}& \ident{number}  \\
        &\narrow{\lsep}& \ident{any}  &\narrow{\lsep}& \ident{collection} \\
        &\narrow{\lsep}& \ident{record} &\narrow{\lsep}& \ident{named}\;~\nu 
\end{array}
%
\quad\;\;\;
%
\begin{array}{rcl}
 \tytagof(\ident{string}) &\narrow{=}& \ident{string}\\
 \tytagof(\ident{bool}) &\narrow{=}& \ident{bool}\\
 \tytagof([\sigma]) &\narrow{=}& \ident{collection}\\
 \tytagof(\hat{\sigma}~\kvd{option}) &\narrow{=}& \tytagof(\hat{\sigma})
\end{array}
%
\;\;
%
\begin{array}{rcl}
 \tytagof(\kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl) &\narrow{=}& \ident{any}\\
 \tytagof(\{ \nu_1 : \sigma_1, \; \ldots \; , \nu_n : \sigma_n \}) &\narrow{=}& \ident{rec-anon}\\
 \tytagof(\nu\; \{ \nu_1 : \sigma_1, \; \ldots \; , \nu_n : \sigma_n \}) &\narrow{=}& \ident{rec-named}~\nu \\
 \multicolumn{3}{c}{ \tytagof(\sigma) = \ident{number}\quad \sigma \in\{ \ident{int},\ident{decimal},\ident{float} \}  } 
\end{array}
%
\\[3em]
%
\begin{array}{rcll}
 \addopt{\hat{\sigma}} &\narrow{=}& \hat{\sigma}~\kvd{option} &(\textnormal{non-nullable types})\\
 \addopt{\sigma} &\narrow{=}& \sigma &(\textnormal{otherwise})
\end{array}
\qquad\qquad
\begin{array}{rcll}
 \dropopt{\hat{\sigma}~\kvd{option}} &\narrow{=}& \hat{\sigma} &(\textnormal{option})\\
 \dropopt{\sigma} &\narrow{=}& \sigma &(\textnormal{otherwise})
\end{array}
\end{array}
}
\end{equation*}
